Simulating High-Dimensional Multivariate Data

### 2021-02-24 040-monte-carlo-evaluation.Rmd
document renders

```{r ch040-LoadLib040, include=FALSE}
library(tidyverse)
library(bigsimr)

@@

cores <- as.integer(parallel::detectCores() - 1)
set.seed(10162020)
```

@@

Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations. We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details). Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3e-04, r_1 = r_2 = 4, \rho \in \Omega$.

@@

| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
|-------------|:--------------:|----------------------:|
|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho_i), i=1,\ldots,100$ |

@@

```{r ch040-biNormPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
dat <- readRDS("./results/biNorm_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()
## ggsave('fig/plot-biNormPlot.pdf')
```

```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
dat <- readRDS("./results/biGamma_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()
## ggsave('fig/plot-biGammaPlot.pdf')
```

```{r ch040-biNegBinPlot, eval=FALSE, fig.path='fig/plot-', dev='png', fig.ext='png', fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
dat <- readRDS("./results/biNB_sims.rds")
dat %>%
    filter(cores == cores) %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    ## ggplot(aes(rho, rho_hat)) +
    geom_point() +
    geom_abline(slope = 1) +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N) + theme_bw()
## ggsave('fig/plot-biNegBinPlot.pdf')
```

```{r ch040-combineBiSims, cache=FALSE}
## compare relative differences in MSE for each dependency, compared to Pearson
allDat <- NULL
dat <- readRDS("./results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
dat <- readRDS("./results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
```

@@

```{r ch040-BiError}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMAE  <- allDat %>%
    ## filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( MAE = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse

## AGS 12/22/20: Removing the latex formatting fixes the html render.
## But how to improve latex table formatting? Better to use the template
knitr::kable(tabMAE,
             ## format = 'latex',
             col.names = c("No. of random vectors",
                           "Correlation type",
                           "Distribution",
                           "Mean abs. error"),
             ## booktabs = TRUE,
             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
             ## linesep = c('', '',  '\\addlinespace')
)
```

@@

```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap = "Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )

# Set colors
## RColorBrewer::display.brewer.all()
numColors <- 4
numGroups <- length(levels(allDat$margins))
## myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Greys')[ ((numColors - numGroups) + 1): numColors  ] )
myColors <- rev( RColorBrewer::brewer.pal(n = numColors, name = 'Blues')[ ((numColors - numGroups) + 1): numColors  ] )

allDat %>%
    ## ggplot(aes(x = rho, y = rho_hat, color = margins, shape = margins)) +
    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
    ## ggplot(aes(rho, rho_hat)) +
    ## geom_point(alpha = 1, size = 2) +
    geom_point(size = 2) +
    ## scale_shape_manual(values=c(1, 2, 3))+
    ## scale_color_manual(values=c('#999999','#E69F00', '#56B4E9')) +
    scale_color_manual(values = myColors ) +
    geom_abline(slope = 1, linetype = 'dashed') +
    labs(x = "Specified Correlation", y = "Estimated Correlation") +
    ## facet_wrap(~ + N) + theme_bw()
    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
    theme_bw() +
    theme(legend.position = "bottom", legend.direction = "horizontal")

## allDat %>%
##     ggplot(aes(x = rho, y = rho_hat, color = type, shape = type)) +
##     ## ggplot(aes(rho, rho_hat)) +
##     geom_point(alpha = 0.75) +
##     geom_abline(slope = 1) +
##     labs(x = "Specified Correlation", y = "Estimated Correlation") + 
##     ## facet_wrap(~ + N) + theme_bw()
##     facet_wrap(~ margins + N) + theme_bw()

## ggsave('fig/plot-biNormPlot.pdf')
```

>>>>

```{r ch040-LoadLib040, include=FALSE}
devtools::load_all()
library(tidyverse)
library(bigsimr)

@@

cores <- as.integer(parallel::detectCores() - 1)
set.seed(10162020)
prob <- 3e-04
```

@@

Table \@ref(tab:sims) lists our identical-marginal, bivariate simulation configurations. We increase the simulate replicates $B$ to check that our results converge to the target correlations and gauge statistical efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see [Example applications](examples) for estimation details). Thus we arrive at a multivariate negative binomial (MVNB) $p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega$.

@@

| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
|-------------|:--------------:|----------------------:|
| $1000$    | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
| $10,000$  | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
| $100,000$ | Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ |

@@

```{r ch040-biNormPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for MVN."}
bivariate_normal_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-biGammaPlot, eval=FALSE, fig.height=5, fig.width=8, fig.cap="`bigsimr` recovers the Pearson specified correlations for Bivariate Gamma."}
bivariate_gamma_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
## ggsave('fig/plot-biGammaPlot.pdf')
```

```{r ch040-biNegBinPlot, eval=FALSE, fig.width=8, fig.height=5, fig.cap="`bigsimr` recovers the correlations for bivariate negative binomial only approximately for Pearson but (nearly) exactly for the rank-based correlations."}
bivariate_nbinom_sims %>%
    ggplot(aes(rho, rho_hat, color = type)) +
    geom_point() +
    geom_abline(slope = 1) +
    facet_wrap(~ type + N) + 
    theme_bw()
```

```{r ch040-combineBiSims, cache=FALSE}
allDat <- bind_rows(
    select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
    select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
    select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
) %>%
    mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
           type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
           N = factor(N))
```

@@

```{r ch040-BiError}
tabMAE  <- allDat %>%
    group_by(N, type, margins) %>%
    summarize(MAE = mean(abs(rho - rho_hat))) %>%
    ungroup()

## AGS 12/22/20: Removing the latex formatting fixes the html render.
## But how to improve latex table formatting? Better to use the template
tabMAE %>%
    knitr::kable(booktabs = TRUE,
                 col.names = c("No. of random vectors",
                               "Correlation type",
                               "Distribution",
                               "Mean abs. error"),
                 caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.')
```

@@

```{r ch040-bPlot, cache=FALSE, fig.asp=1.30, fig.weight=4, fig.cap="Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c("Pearson", "Spearman", "Kendall")
names(typeLabs) <- c("Pearson", "Spearman", "Kendall")

# Set colors
## RColorBrewer::display.brewer.all()
numColors <- 4
numGroups <- length(levels(allDat$margins))
myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Blues")[ ((numColors - numGroups) + 1):numColors])

allDat %>%
    ggplot(aes(x = rho, y = rho_hat, color = margins)) +
    geom_point(size = 2) +
    scale_color_manual(values = myColors ) +
    geom_abline(slope = 1, linetype = "dashed") +
    labs(x = "Specified Correlation", y = "Estimated Correlation") +
    facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
    theme_bw() +
    theme(legend.position = "bottom", legend.direction = "horizontal")
```
